---
title: "The Relationship of SES and Reading Scores of High School Seniors"
author: "Mitchell Toomey"
output: html_notebook
---

```{r echo=FALSE}

# The "here" library will let me read the data from an individual folder.

library(here)

# I read the data and put it in an object called "students".

students <- read.csv(here("Data", "hsb2.csv"), header = TRUE, sep = ",")

# SES and race are categorical variables, what we call "factors," so I need to
# set them as factors.

students$ses <- factor(students$ses,
                       labels = c("Low", "Middle", "High"))

students$race <- factor(students$race,
                        labels = c("Hispanic", "Asian", "African-American",
                                   "White"))

```

### Description

This report will study the relationship of socioeconomic status (SES) and
reading test scores of high school seniors. The data for this study is from a
randomly selected national sample of 200 high school seniors. First, I will
study the composition of SES in the sample. I will then address the main
purpose of this study by comparing reading scores for different SES groups.
Finally, I will see how SES is related to ethnic groups as a means of
potentially explaining why reading scores often are found to differ among
ethnic groups, as shown in past studies.

*Source of data: The "High School and Beyond" survey conducted on high school*
*seniors in 1980 by the National Center of Education Statistics.*

### Analysis

In this data set, SES is divided into three categories: low, medium, and high.
I do not know the definitions the original researchers used for these
classifications. Figure 1 illustrates the composition of SES in the sample.

\newpage

***

*Figure 1. Bar plots with proportion of sample of 200 high school seniors in*
*SES classifications.*

```{r echo=FALSE}

# I calculated the frequencies of students in SES categories.

ses.counts <- table(students$ses)

# I calculated the proportions of students in SES categories.

ses.props <- prop.table(ses.counts)

# I constructed a bar graph using these proportions.

barplot(ses.props,
        xlab = "SES Groups",
        ylab = "Proportion", ylim = c(0, 0.5))

```

***

```{r echo=FALSE}

# I put the frequencies and proportions in a table, but there are only a few
# of these so I put those in a narrative instead of showing the table.

cbind(ses.counts, ses.props)

```

The sample consists of 23.5% in the low SES category, 47.5% in the middle SES
category, and 29.0% in the high SES category.

Figure 2 illustrates how reading test scores are related to SES categories.
From this we can see that the central tendency of the reading scores in the
high SES category outperforms that of the middle SES category, which in turn
outperforms the low SES category. Additionally, the variation is similar for
each of the categories.

\newpage

***

*Figure 2. Side-by-side boxplots illustrating reading test scores for SES*
*categories.*

```{r echo=FALSE}

# This command creates side-by-side boxplots.

boxplot(students$read ~ students$ses,
        ylim = c(20, 80),
        xlab = "SES Level",
        ylab = "Reading Score")

```

***

Table 1 is a table of descriptive statistics for reading test scores in each of
the three SES categories. For every measure of location (i.e. every statistic
in the table except the standard deviation), the change to a higher SES is
associated with an increase in the statistic for the reading test scores. That
is, the change in SES is associated with a change in reading test scores across
the range of test scores. For example, the mean for the middle SES students is
about 3 points higher than for low SES students. Similarly, the mean for the
students in the high SES group is about 5 points higher than for the middle
SES group.

***

*Table 1. Statistics for each of the three SES categories.*

```{r echo=FALSE}

# I calculated descriptive statistics for reading scores of SES groups.

Mean <- tapply(students$read, students$ses, mean)
SD <- tapply(students$read, students$ses, sd)
Min <- tapply(students$read, students$ses, min)
Q1 <- tapply(students$read, students$ses, quantile, p = 0.25)
Med <-  tapply(students$read, students$ses, median)
Q3 <- tapply(students$read, students$ses, quantile, p = 0.75)
Max <- tapply(students$read, students$ses, max)

# I then created a "column bind" to put these in a table. The "round" function
# eliminates the decimal points to provide numbers easier to read and remember.

round(cbind(Mean, SD, Min, Q1, Med, Q3, Max))

```
***

The sample data also includes four ethnic categories: Hispanic, Asian, African
American, and white. The relationship of SES to ethnic group is illustrated
in Figure 3.

***

*Figure 3. Bar plots with percent of each SES category within ethnic groups.*

```{r echo=FALSE}

# This constructs a two-way table of frequencies.

ses.race.counts = table(students$ses, students$race)

# This constructs conditional percentages (conditioned on ethnic group) of
# SES categories.

ses.race.percents = 100*prop.table(ses.race.counts, 2)

# This provides a stacked bar graph.

barplot(ses.race.percents,
        xlab = "Ethnic Group",
        ylab = "Percents",
        legend = c("Low", "Middle", "High"),
        args.legend = list(title = "SES", cex = .7))

```

***

Table 2 provides the percentages of each SES within each ethnic group. It is
apparent that there is a relationship of SES and ethnicity. A relatively low
percentage of whites are in the low SES category, whereas the percentages of
non-white ethnic groups in the low SES categories are substantially higher than
for whites; in fact, it is three times higher for African Americans.

***

*Table 2. Percent of each SES classification within each of four ethnic*
*groups.*

```{r echo=FALSE}

# This provides percents. The "round" function eliminates decimal points to
# make the numbers easier to read and remember.

round(ses.race.percents)

```
***

### Conclusions

I examined whether reading test scores are related to socioeconomic
classification. The answer is “yes.” As socioeconomic status increases, the
reading test scores also increase. This is true across the range of the reading
test score spectrum. For example, the mean for the middle SES students is
about 3 points higher than for low SES students. Similarly, the mean for the
students in the high SES group is about 5 points higher than for the middle
SES group. This upward shift has very little influence on the width of the
distribution of scores, so it is an “across the board” shift.

As a secondary question, I also looked at the relationship of SES to ethnic
groups, finding that there is a clear relationship. A substantially higher
percentage of white students come from higher SES homes than their Hispanic,
Asian, and, most especially, African American counterparts. As an example,
the percentage of African American high school seniors from low SES homes is
more than three times the percentage of white seniors from low SES homes. This
relationship may help explain differences in reading scores for various ethnic
groups, though that was not the focus of this report and more study will be
needed to confirm that this is the case.

### Press Release

Researchers at the University of South Carolina used a national sample of 200
high school seniors to study the relationship of socioeconomic status to
reading test scores. “These findings are important,” stated lead analyst Mike
Seaman, “because SES widely varies and could be related to school resources and
instructional environment.” Seaman said that about half of the seniors are from
middle-income homes, with the other half about evenly split between high- and
low-income homes. “That’s not a complete picture,” he said, “because the
distribution is very different for different ethnic groups. For example, the
percentage of African American seniors in low-income homes is over three times
the percentage of whites in low-income homes.”

The main findings of the study show that there is indeed a relationship of SES
to reading test scores, with test scores increasing as income-level increases.
Seaman reports the difference in average reading test scores for seniors from
a high SES environment is about 8 points higher than that of seniors from a low
SES home.
