---
title: "Exploring Loops"
author: "Michael A. Seaman"
date: "24 April 2020"
version: 1.0
output: html_notebook
---

```{r include=FALSE}

library(tidyverse)
library(here)

```


This is a presentation about loops and how to eliminate them.

First, let us look at a loop and why we use them. I'm going to create a
function that will calculate the mean of a vector of numbers.

```{r}

my_mean <- function(y) {

  total_sum <- 0
    
  for (i in 1:length(y)) {
    total_sum <- total_sum + y[i]
  }
  
  return(total_sum / length(y))
}

```


Call the function.

```{r}

my_numbers <- c(1, 3, 5, 7, 9)

my_mean(my_numbers)

```

It would be easier, of course, to just use the mean function in Base R. This
function is vectorized. That means that behind the scenes it uses a loop, but
the loop has been compiled to run very fast. Vectorized functions allow us to
use compiled loops without having to write loops ourselves.

```{r}

mean(my_numbers)

```

Even with vectorized functions, it sometimes would appear that we still need
a loop. Consider the following matrix.

```{r}

my_matrix <- matrix(c(14, 21, 33, 42,
                      19, 31, 28, 55,
                      8,  27, 28, 48,
                      11, 26, 30, 51),
                    ncol = 4,
                    byrow = TRUE)

my_matrix

```

Suppose that we developed a new statistic based on squaring the difference 
between the mean and the median. Squaring, of course, helps us look at 
the size of differences regardless of whether the mean is above or below the median. This statistic will be small for symmetrical distributions, but get larger for skewed distributions. Here's a function for our new statistic.

```{r}

mean_med_sqr <- function(y) {

    return((mean(y) - median(y))^2)
}

```

Now I want to apply this function to the first column of my matrix. Here's
the first column.

```{r}

my_matrix[, 1]

```

Now here's what I get when I calculate my new statistic.

```{r}

mean_med_sqr(my_matrix[, 1])

```

So far, so good! Now suppose I want to apply this function to each of the
columns of my matrix. There are only four of them, so I could do it like this.

```{r}

mean_med_sqr(my_matrix[, 1])
mean_med_sqr(my_matrix[, 2])
mean_med_sqr(my_matrix[, 3])
mean_med_sqr(my_matrix[, 4])

```

That works, but what if I had 100 columns in my matrix. Yikes! That's a lot of
cutting and pasting and changing. Instead, I could use a for loop. For loops 
are a bit like functions because they hide what is going on inside them, so to
see the results I need a print statement.

```{r}

for (i in 1:ncol(my_matrix)) {
  print(mean_med_sqr(my_matrix[, i]))
}

```

OK, that's cool and it would work whether I have four columns or 400 columns.
Yet I used a for loop and this video series is about getting rid of them.
So, let's get rid of it! The apply function in Base R will apply our function 
to every column. So we are already using vectorized functions (mean, median,
and squaring), but now we will vectorize across the columns of our matrix as 
well. This is vectorizing our vectorizing!

You'll probably wonder what the number "2" is doing in this. It is telling R
to vectorize across the second dimension (the columns). If we used a "1" it
would apply the function for every row instead.

```{r}

apply(my_matrix, 2, mean_med_sqr)

```

The apply function only works with matrices and arrays. As you know, we usually read our data into a data frame. Let's put the same data in a data frame.

```{r}

my_df <- data.frame(y1 = c(14, 19, 8, 11),
                    y2 = c(21, 31, 27, 26),
                    y3 = c(8, 27, 28, 48),
                    y4 = c(11, 26, 30, 51))

my_df

```

We can now see our four variables and we want to apply our function to all 
four. We cannot use apply, because this is a data frame, but we can use lapply!
The "l" in lapply stands for "list." This function will work with any list.
A data frame, of course, is just a special list with equal-length vectors,
so we can use lapply. We could also use lapply with a list (of course!).

```{r}

lapply(my_df, mean_med_sqr)

```

We got what we wanted without a loop! Our one concern might be that the output 
is also in a list. We probably would prefer a vector. This is where sapply 
comes to the rescue! It does the same job as lapply, but returns the results
in a more simple vector. (Maybe it is named sapply because it is "simple" 
apply.)

```{r}

sapply(my_df, mean_med_sqr)

```

Finally, consider the hsb2 data.

```{r}

hsb2 <- read_csv(here("Data", "hsb2.csv"))

```

What if we wanted to use our function on math scores by SES? We could separate 
out the three different levels of SES into three data sets, then look at the 
square of the mean and median difference of math scores in all three data sets. Suppose, however, we had a different categorical variable that had 20  categories. That would be a pain! Then we might be tempted to use a for loop 
to separate out the different categories and another for loop to go through 
each of the 20 data sets to calculate this statistic. That would probably be 
easier than separating all 20 data sets and then doing the mean for all 20 data sets by hand. This is easier yet!

```{r}

tapply(hsb2$math, hsb2$ses, FUN = mean_med_sqr)

```

The tapply function vectorizes our function not only for the variable of 
interest (math), but also for the grouping variable (ses). Two vectorizations 
for the price of one!

So these are the methods of asking for a vectorization in Base R and thus 
avoiding loops: apply, lapply, sapply, and tapply.
