---
title: "R Data Frames"
author: "Michael A. Seaman"
output: html_notebook
---

```{r include=FALSE}

library(here)

```


The data frame is the most common data structure for storing research data.
You can think of a data frame as a spreadsheet. Each column is an atomic
vector, but different columns can contain different types of data. Thus, a
data frame is a special type of list in which each element (i.e., each
column in our spreadsheet) is equal length.


### Creating Data Frames

We can create data frames by entering the data within our script, but it is
far more common to read in the data using one of the read functions. You can
also create data frames by binding vectors together into columns or rows.

```{r include=FALSE}

# We can use the data.frame function to create a data frame.

my_data <- data.frame("ID" = 1:4,
                      "Name" = c("Greg", "Sara", "Manuel", "Liz"),
                      "Score" = c(23, 27, 26, 29))

# Let's look at the characteristics of this data frame.

class(my_data)
attributes(my_data)
str(my_data)

# Data frame columns and rows must have names, so if you do not name them,
# R will name them for you and it's not pretty!

my_data <- data.frame(1:4,
                      c("Greg", "Sara", "Manuel", "Liz"),
                      c(23, 27, 26, 29))

# Type this in the console:

# View(my_data)

# In some versions of R, character vectors are converted to factors. We can
# suppress this with the stringsAsFactors argument.

my_data <- data.frame("ID" = 1:4,
                      "Name" = c("Greg", "Sara", "Manuel", "Liz"),
                      "Score" = c(23, 27, 26, 29),
                      stringsAsFactors = FALSE)

str(my_data)

# We could also create the vectors first and then combine into a data frame.

ID <- 1:4
Name <- c("Greg", "Sara", "Manuel", "Liz")
Score <- c(23, 27, 26, 29)

my_data <- data.frame(ID, Name, Score)

str(my_data)

# Another way to preserve the character vector, we can use the I() function.
# In this context, "I" stands for "insulate". This prevents the character
# vector from being changed to a factor both now and in any future functions.

ID <- 1:4
Name <- c("Greg", "Sara", "Manuel", "Liz")
Score <- c(23, 27, 26, 29)

my_data <- data.frame(ID, I(Name), Score)

str(my_data)

# Note that Name is now an "AsIs" character vector. This is part of the
# object to tell any functions to leave this character vector as it is.

class(my_data$Name)

# The most common way to create a data frame is by reading it in from
# an external file.

Stat_Data <- read.csv(here("Data", "Statistics Class Data.csv"))

str(Stat_Data)

# To preserve character vectors with read.csv, regardless of R version, we can
# again set the stringsAsFactors argument to FALSE.

Stat_Data <- read.csv(here("Data", "Statistics Class Data.csv"),
                      stringsAsFactors = FALSE)

str(Stat_Data)

# Columns can be added to a data frame by adding a named vector.

my_data$Gender <- c("Male", "Female", "Male", "Female")

# We can also add columns by putting them into a data frame and then
# bind the two data frames together.

new_vars <- data.frame("Age" = c(14, 13, 14, 14),
                       "Transfer" = c(TRUE, FALSE, FALSE, FALSE))

my_data <- cbind(my_data, new_vars)

my_data

# If we want to add rows to a data frame, we need to read those rows
# into a data frame and then bind the two data frames together.

new_data <- data.frame("ID" = 5:6,
                       "Name" = c("Laura", "Patricia"),
                       "Score" = c(28, 25),
                       "Gender" = c("Female", "Female"),
                       "Age" = c(14, 13),
                       "Transfer" = c(TRUE, FALSE))

my_data <- rbind(my_data, new_data)

my_data

# If we want to change column names, we can do this with the colnames
# function.

colnames(my_data) <- c("id", "name", "score", "gender", "age", "transfer")

my_data

# We can also use subsetting to change certain column names.

colnames(my_data)[2] <- "first.name"

my_data

# We can change row names with the row.names function.

row.names(my_data) <- c(101:106)

my_data

```


### Subsetting Data Frames

A data frame is a special type of list, so the methods that we learned for
subsetting and extracting with a list work with a data frame. Additionally,
because a data frame is rectangular, like a matrix, some of the matrix subset
methods work as well.

```{r}

# Data frames can be subset like lists and matrices. Here's one column
# obtain through subsetting multiple ways.

my_data[,3]
my_data[[3]]
my_data$score

# Here are two rows.

my_data[2:3,]

# Here are all the rows except the 2nd one.

my_data[-2,]

# Here are three columns and two rows.

my_data[c(3,4), 2:4]

# You can also subset with the subset function and a logical Boolean operator.

my_data_females <- subset(my_data, gender == "Female")

my_data_females

```


### Reviewing Data Frames

Making certain that data frames contain exactly what they should and are in
the structure that they should be in prior to analysis is an important step in
the research process. We will look at a few of the functions that can assist
with this process.

```{r include=FALSE}

# We already know about the structure function.

str(Stat_Data)

# We can summarize the data.

summary(Stat_Data)

summary(my_data)

# We can make sure the data appears to be set up correctly by looking at
# the first few elements.

head(Stat_Data)

# Or the last few.

tail(Stat_Data)

# Of course if the data set is not too large, we might want to look at the
# entire thing! We can do this with the View() function. Try it!

# View(Stat_Data)

# The nrow and ncol functions will report the number of rows and columns.

nrow(Stat_Data)
ncol(Stat_Data)

# Or we can use dim.

dim(Stat_Data)

```


### Common Data Frame Functions

As with lists, there are a huge number of functions to use with data frames.
We will look at a small sampling of these.

```{r}

# Like matrices, data frames can be transposed.

my_data
t(my_data)

# The aggregate function allows you to apply a function to subsets in your
# data frame. Let's look at mean scores for males and females.

aggregate(my_data[c("score", "age")], by = list(my_data$gender), FUN = mean)

# Note that aggregate works on data frames and returns a data frame.

my_data_means <- aggregate(my_data[c("score", "age")],
                           by = list(my_data$gender),
                           FUN = mean)

my_data_means

class(my_data_means)

# Could I have used aggregate for the entire data set? Yes, if I don't
# care about getting a bunch of warning messages for non-numeric data.

aggregate(my_data, by = list(my_data$gender), FUN = mean)

# Let's use the HSB2 data for a more complex example.

hsb2 <- read.csv(here("Data", "hsb2.csv"))

# Let's obtain test score means by both gender and type of school.

aggregate(hsb2[7:11], by = list(hsb2$female, hsb2$schtyp), FUN = mean)

# It doesn't have to be just for means!

aggregate(hsb2[7:11], by = list(hsb2$female, hsb2$schtyp), FUN = fivenum)

# The apply function will apply another function, provided as an argument
# to the data frame.

apply(hsb2, 2, median)

# The apply function returns a vector, rather than a data frame.

hsb2_medians <- apply(hsb2, 2, median)

hsb2_medians

class(hsb2_medians)

# The "2" given as the second argument specified columns. A "1" would
# give us row information.

apply(hsb2, 1, median)

# The lapply function is designed to work on lists, but will also work for
# data frames because a data frame is a list of vectors of equal length.
# Since it expects to apply the function to the elements of a list, we don't
# need to specify whether to use columns or rows. It will only do this for
# columns.

lapply(hsb2, median)

# Note that it returns a list.

hsb2_median_list <- lapply(hsb2, median)

class(hsb2_median_list)

# If we want a vector instead, we can use sapply. This is like using apply,
# except assuming a list, so no need to specify that we want to perform
# the function on columns.

hsb2_median_vector <- sapply(hsb2, median)

hsb2_median_vector

class(hsb2_median_vector)

# We can dig deeper and break the analysis into pieces with tapply.
# Let's suppose we want medians for reading by SES.

tapply(hsb2$read, hsb2$ses, median)

# If you want to cross tabulate for more than one variable, you can do it
# using a list. Let's obtain medians for reading by SES and school type.

tapply(hsb2$read, list(hsb2$schtyp, hsb2$ses), median)

# Here's a function that you should avoid, but I want you to be aware that
# it exists so that when you hear about it, you'll know to avoid it.

attach(my_data)

# Now everything we do assumes we are using my.data.

first.name
first.name[3:4]

cbind(first.name, score)

mean(score)

# Let see why we should avoid attach. First, let's detach. 

detach(my_data)

# Now let's create a vector called score and calculate the mean.

score <- c(125, 315, 282)
mean(score)

# Let's reattach my.data and calculate the mean of the score in my.data.

attach(my_data)

first.name
mean(score)

# R even warned us about this. We might think we are accessing score within
# the my_data data frame, but we are not. This can lead to serious problems
# if you are analyzing data to prepare a manuscript! If you look at the
# Google R Style Manual, Google tells employees not to use attach. We won't
# either. We need to end this potentially destructive attachment.

detach(my_data)

# There are alternatives. We already know that one alternative is to directly
# reference a variable.

mean(hsb2$read)

# Another is to use the data frame as an argument, which is permitted by
# some functions.

lm(write ~ read, data = hsb2)

# A third is to use the with function.

with(hsb2, mean(read))

with(hsb2, lm(write ~ read))

```
