---
title: "R Functions"
author: "Michael A. Seaman"
output: html_notebook
---

```{r include=FALSE}

library(here)

```


Functions are the workers in R. The purpose of a function is to perform a
task that we are going to want to perform multiple times.


### Basic Components of Functions

Functions consist of a name, arguments, and the body of the function (which we
cannot see when we call a function). Functions return an object of some
class, but we often use the object directly rather than store it.

```{r}

# We will use the mean function as an example. We can learn about the mean
# in the help section of R or by doing a web search on "R mean". The name
# of the function is "mean".

?mean

# The mean is a function.

class(mean)

# Given that functions are objects, we can assign functions to other objects.

super_duper_mean <- mean

class(super_duper_mean)

# There is one required argument. We must provide an object. The mean will
# work on various types of objects. (Examples of this come later.)

my_data <- c(2, 2, 3, 3, 4, 5, 7, 7, 9, 11)
class(my_data)

mean(my_data)
super_duper_mean(my_data)

# There are two optional arguments for the mean: trim and na.rm.

# Let's use the mean function again, but this time trim the lowest and
# highest 10% of the values.

mean(my_data, 0.1)

# Let's check what this is doing.

mean(c(2, 3, 3, 4, 5, 7, 7, 9))

# We can use argument names when we run the function. This is never a bad
# idea because it helps document our code.

mean(x = my_data, trim = 0.1)

# This next example is not great practice, but it works. By convention, we
# call mandatory arguments first, before optional arguments. As this example
# illustrates, this is not required.

mean(trim = 0.1, x = my_data)

# Suppose we have some missing data.

my_next_data <- c(2, 2, NA, 3, NA, 5, 7, 7, 9, 11)

mean(my_next_data)

# R tells us that we have some missing data. Here's how to solve the problem.

mean(my_next_data, na.rm = TRUE)

# documentation states that we can pass extra arguments that are not  needed
# for the mean. (Note the ellipses or "three dots" in the documentation.) Let's
# see what happens when we do that.

mean(my_data, trim = 0, na.rm = TRUE, 7, FALSE)

# Nothing happens! So why have three dots? Why allow extra arguments? Good
# question! Let's explore this by looking at the body of the mean function.

body(mean)

# The mean function is very short. It calls the "UseMethod" function with a
# "mean" argument. That tells us that the mean function has methods.

methods(mean)

# There are various functions that our mean function can call, depending on
# the type of object we send to mean. These are called "methods. The
# "UseMethod" function will look at the type of object we sent and then can
# determine which mean function to call. When our object is either numeric or
# logical, it will call mean.default.

body(mean.default)

# To look at using other methods, let's set up some dates. R has a date object!

Valentines_Day <- as.Date("2021-2-14")
St_Patricks_Day <- as.Date("2021-3-17")

both_dates <- c(Valentines_Day, St_Patricks_Day)

# Let's see what happens if we send both dates to the mean function.

mean(both_dates)

# The function "mean" has a method for dates that finds the middle date
# between two dates. Cool! There are other methods for the mean, but we will
# stop here after having looked at the default method and also a method for
# working with dates. Now back to the three dots. Let's look at the arguments
# listed for the mean.

args(mean)

# What? Only an object and the three dots? Yes! Remember that when we call the
# mean function that it is checking what type of object we have before
# deciding which mean method to use. If we send a numeric object, then it will
# call mean.default.

args(mean.default)

# This function has three possible arguments. Thus, the three dots in the
# original call to the mean allows for additional arguments, just in case the
# method we end up calling has more arguments. Eureka! Three dots has (high)
# value!

# The final component of a function is what it returns.

my_data_mean <- mean(my_data)

class(my_data_mean)

```


### A Look at a Few Functions

Let's see the usefulness of functions by looking at a few of them.

```{r}

MountJuniper_Data <- read.csv(here("Data",
                                   "Mount Juniper Elementary Data.csv"),
                              header = TRUE,
                              sep = ",")

# Let's take out names for security purposes

New_MJ_Data <- MountJuniper_Data

New_MJ_Data$First.Name <- NULL
New_MJ_Data$Last.Name <- NULL

# Let's also set grade level as a factor

New_MJ_Data$Grade <- factor(New_MJ_Data$Grade)

# Now let's save to a CSV file and an RData file.

write.csv(New_MJ_Data,
          here("Data", "Mount Juniper Elementary Data (no names).csv"))

# Not quite what we wanted! Let's look at the arguments for this function.

?write.csv

write.csv(New_MJ_Data,
          here("Data", "Mount Juniper Elementary Data (no names).csv"),
          row.names = FALSE)

save(New_MJ_Data,
     file = here("Data", "Mount Juniper Elementary Data (no names).RData"))

# Let's see why the "file =" is needed.

?save

# Let's look at our objects, clear the environment and load the data.

ls()
rm(list = ls())
ls()

load(here("Data", "Mount Juniper Elementary Data (no names).RData"))

# Now let's look at side-by-side box plots for the scores, by grade.

boxplot(New_MJ_Data$Score ~ New_MJ_Data$Grade)

# We need axes labels. There are many additional arguments that we can
# use to spiff this up.

boxplot(New_MJ_Data$Score ~ New_MJ_Data$Grade,
        names = c("Third", "Fourth", "Fifth"),
        xlab = "Grade Level",
        ylab = "Score",
        main = "Scores at Mount Juniper Elementary by Grade Level",
        border = "coral4",
        col = c("cyan", "darkolivegreen4", "burlywood1"))

```

