---
title: "R Lists"
author: "Michael A. Seaman"
output: html_notebook
---

```{r include=FALSE}

library(here)

```


A list is a vector of vectors that can be of different types and different
lengths. Databases and research data sets almost always contain various types
of data, so lists (and data frames, which are a special type of list) are
the most common data structure for storing the set of data, as a whole. Even
though we may take individual vectors out of the list for analysis, knowing
how to create and then access elements of a list is vitally important to
using R for research.


### Creating Lists

We can create lists by entering list data or by combining vectors. The vectors
can be of different types and length. This makes lists ideal for storing
output of functions, such as hypothesis testing functions, analysis of
variance functions, linear models, and so on.

```{r}

# We can use the list function to create a list.

our_course <- list("EDRM 718",
                   c("Tuesday", "Thursday"),
                   c("presentations", "writing scripts", "learning groups"),
                   c(1:17),
                   list(c("homework", "checkpoints", "exam"),
                        c(50, 20, 30)))

our_course

# We can also do this with existing vectors.

course_num <- "EDRM718"
meeting_times <- c("Tuesday", "Thursday")
activities <- c("presentations", "writing scripts", "learning groups")
student_id <- 1:17
graded_elements <- list(c("homework", "checkpoints", "exam"),
                        c(50, 20, 30))

our_course <- list(course_num,
                   meeting_times,
                   activities,
                   student_id,
                   graded_elements)

our_course

# List elements can be named with the names function.

names(our_course) <- c("title",
                       "days",
                       "activities",
                       "id",
                       "grading")

our_course

# We can also name elements when we create the list.

our_course <- list("title" = "EDRM 718",
                   "days" = c("Tuesday",
                              "Thursday"),
                   "activities" = c("presentations",
                                    "writing scripts",
                                    "learning groups"),
                   "id" = c(1:17),
                   "grading" = list(c("homework", "checkpoints", "exam"),
                                    c(50, 20, 30)))

our_course

# Let's look at the class and attributes.

class(our_course)
attributes(our_course)

# Note that the attributes are themselves a list.

our_course_attributes <- attributes(our_course)

class(our_course_attributes)


```


### Subsetting and Extracting from Lists

With the classes of objects that we have studied up to now, subsetting and
extracting was the same thing. Not so with lists.

```{r}

# Elements of a list can be accessed by order number, just like vectors.

our_course[1]

our_course[3]

our_course[5]

# Not that what is actually being returned is a subset of a list, which
# is also a list.

our_activities <- our_course[3]

our_activities

class(our_activities)
typeof(our_activities)

# If we want to pull out a vector from the list, we can use the extraction
# operator [[]]. This operator "digs into the list" and looks at content.

our_activities <- our_course[[3]]

our_activities

class(our_activities)
typeof(our_activities)

# If the elements of the list are named, the $ operator serves the same
# purpose as the extraction operator. The advantage of $ is brevity and
# simplicity. The disadvantage is that the elements of the list must be
# called by name. For example, you couldn't use $ in a for loop.

our_activities <- our_course$activities

our_activities

class(our_activities)
typeof(our_activities)

# We can also use names with the [[]] operator.

our_course[["activities"]]

# Note the use of the extraction operator to obtain an element from our
# list within a list.

grading_percents <- our_course[[5]][[2]]

grading_percents

# Or alternatively like this.

grading_percents <- our_course$grading[[2]]

grading_percents

# That raises a question. Can we name elements of a list within a list? Yes!

names(our_course[[5]]) <- c("requirements", "percents")

our_course$grading$requirements
our_course$grading$percents

# As with vectors, we can leave off a list item.

our_course[-4]

# Why doesn't this next one work? (Try typing it in the console.)

# our_course[[-4]]

# But this does!

our_course[[4]][-3:-5]

# We can use logicals to subset a list.

our_course[c(TRUE, TRUE, FALSE, FALSE, TRUE)]

```

### Modifying a List

We can modify elements of a list much like we can modify elements of other
object classes. Keep in mind, however, that when we are modifying an element
in a vector within the list, we need to modify using extraction operators.

```{r}

# We can modify by assigning new values, just as with a vector.

our_course[1] <- "Research and R"

our_course

our_course$title <- "EDRM 718"

our_course

# Here we are changing a single element within a sub-list.

our_course$grading$requirements[3] <- "final exam"

our_course

# We can add elements to a list.

our_course$instructor <- "Mike Seaman"
our_course$some_students <- c("Silvia", "Allison", "Hyunsu", "Lisa")

our_course

# We can delete list elements.

our_course$instructor <- NULL

our_course

# Note that if we delete an element in the middle, elements below it assume
# a new order in the list.

our_course[5]

our_course$id <- NULL

our_course

our_course[4]

# We can merge lists.

course_info <- list(semester = "Spring 2019",
                    credits = 3,
                    prerequisites = "Two semesters of statistics",
                    variable.credit = FALSE)

our_course <- c(our_course, course_info)

our_course

```


### Basic List Functions

Given the importance of lists to using R for conducting research, there are
many, many functions for working with lists in Base R, and oodles more when
we install packages. Here we will look at just a couple of functions to get a
flavor for how functions with lists might work.

```{r}

# A list can be converted to a vector. Remember that vectors are atomic,
# so various data types in the list will have to be coerced to other types.

course_vector <- unlist(our_course)

course_vector

# We can view the structure of a list with the str function.

str(our_course)

# Data frames can be split into lists with the split function.

# R has quite a few datasets built in. The creators of R must have realized
# that statistics teachers would appreciate this! We can see the directory
# of datasets. Try typing this function into the console.

# data()

# Let's use mtcars. Type this function into the console.

# View(mtcars)

# With the split function we can split a data frame into multiple data
# frames that we store in a list. Let's split mtcars by the number of
# cylinders in the engine.

car_data <- split(mtcars, mtcars$cyl)

str(car_data)

# The lapply function let's us apply a function to elements of a list.

lapply(car_data, function(x) mean(x$mpg))

# We had to use this format because we can't take the mean of every element
# in our list, such as character vectors. If we are using a function that
# can be applied to every element in our list, we can simplify.

lapply(car_data, attributes)

```
