---
title: "R User-Defined Functions"
author: "Michael A. Seaman"
output: html_notebook
---

R functions provide the work in R. They are the doers. Though there are
thousands of functions to choose from, this is still not enough. Sometimes
there is a section of code that we will use over and over. Such a code section
is a good candidate to become a function. By developing our own user-defined
functions, we can add to our function library and increase the power of our
implementation of R in a way that meets our needs.

```{r}

library(here)

```


### Constructing Functions

To create our own function, we need a name, arguments, body, and a return
value.

```{r}

# Here is an example of a function. This is a silly function that squares
# a number and adds five.

square_plus_five <- function (x) {

  our_result <- x^2 + 5
  
  return(our_result)
}

# Now that our function has been defined, let's try calling it.

square_plus_five(7)

# The function is already vectorized!

my_vector <- c(3, 7, 11, 21)

square_plus_five(my_vector)

# I believe that an explicit return() function helps the readability of
# code so that you should use it, but it isn't necessary. If there is no
# return function, R will return the last element that is evaluated in
# the function.

square_plus_five <- function (x) {
  x^2 + 5
}

square_plus_five(7)

# Let's make another function that is a little more general. For this one
# we send it a vector and also tell it how much to add after squaring.

square_plus <- function (x, y) {
  
  our_result <- x^2 + y
  
  return(our_result)
}

# Let's send our vector again, but this time we can add 15 with our new
# function. Notice that we can assign the result to an object.

my_funny_stuff <- square_plus(my_vector, 15)
my_funny_stuff

# What if we only sent one value to our new function. Try typing this:

# my_funny_stuff <- square_plus(my_vector)

# As the error message states, we are missing an argument that does not
# have a default. Let's set one up so that if no second argument is
# included, it will not add anything.

square_plus <- function (x, y = 0) {
  
  our_result <- x^2 + y
  
  return(our_result)
}

# Now this function call works!

my_funny_stuff <- square_plus(my_vector)
my_funny_stuff

my_funny_stuff <- square_plus(my_vector, 15)
my_funny_stuff

# What if we give it too many arguments? Try typing in this next line.

# my_funny_stuff <- square_plus(my_vector, 15, 45)

# That's a problem. If there is a reason to pass through additional arguments,
# we can use the three dots.

square_plus <- function (x, y = 0, ...) {
  
  our_result <- x^2 + y
  
  return(our_result)
}

# Now this works. Note that the 45 is totally ignored.

my_funny_stuff <- square_plus(my_vector, 15, 45)
my_funny_stuff

# So why would we want to include the three dots? We do this if there is any
# chance that inside our function we might want to do something with another
# function. Let's look at an example. Here's a function that will create a
# scatterplot and return the correlation. Let's include three dots.

corr_plot <- function (x, y, ...){
  
  plot(x, y, ...)
  
  corr_xy <- cor(x, y)
  
  return(corr_xy)
}

# Now let's create two vectors and call this function.

explan_var <- c(15, 21, 17, 19, 32, 23, 25, 29, 20, 21)
respon_var <- c(90, 92, 88, 84, 110, 99, 105, 104, 98, 111)

corr_plot(explan_var, respon_var)


# Suppose I decide to send some additional arguments for the plot
# that is inside the function. I can do it!

corr_plot(explan_var,
          respon_var,
          xlab = "Explanatory Variable",
          ylab = "Response Variable",
          col = "red")

```


### Environments

An environment is a space where everything is stored. Normally we are working
in the global environment. There are other environments, including the
environments of packages that contain functions we may not be using in our
current global environment. Functions also create environments.

```{r}

# Let's look at the environment of a function. Most variables are in the
# global environment, such as these variables.

my_var1 <- 20
my_var2 <- 30

# We can look at the environment.

environment()

# Now let's create a function.

add_15 <- function (x) {
    y <- x + 15
  
# While I'm in this function I'm going to set variables with the same
# names as those above and print them.
  
    my_var1 <- 100
    my_var2 <- 200
  
    print(my_var1)
    print(my_var2)
  
# And let's look at the environment while we are here.
  
    print(environment())
  
    return(y)
}

# Now let's call that function.

my_new_num <- add_15(50)
my_new_num

# Let's see the value of our two variables.

my_var1
my_var2

# When a variable name is used, R will first look in the current environment.
# If it does not find it, R will look at the next level up, and so on.

triple_add_five <- function (x) {
  
  # Let's include two nonsense variables in this function.
  
  a <- 10
  b <- 20
  
  # Now we'll create a function within a function that triples a value.
  
  triple <- function (y) {
    
    # We will reset one of our nonsense variables.
    
    a <- 100
    
    # Let's print both a and b.
    
    print(a)
    print(b)
    
    z <- 3*y
    return(z)
  }
  
  # Now let's print a and b again.
  
  print(a)
  print(b)
  
  z <- triple(x) + 5
  return(z)
  
}

# Let's call this function to triple and add 5 to the number 50. Notice that
# the triple function is not in our environment pane. That's because we are
# looking at the global environment.

triple_add_five(50)

```


### Using Functions

We write functions when we want to have a script available for use over and
over again without having to copy in the entire script each time. The examples
above were kept simple (and silly) to introduce the concept of functions. Now
it is time to look at a more useful, but necessarily more complex, user-defined
function.

```{r}

# Here's a bit more complex (and practical!) function just to show what
# a longer function might look like. This function will calculate the
# power or sample size for a one-sided binomial test.

# The arguments are a null hypothesized value, an alternative hypothesized
# value, sample size or power, an indicator of whether this is sample size
# or power, and a confidence level.

binom_up_pow <- function(null_value,
                         alt_value,
                         samp_pow,
                         samp_size = TRUE,
                         conf_lev = 0.95) {

  # Here is a function within the function to calculate power from sample
  # size.
  
  get_power <- function(nv,av,sp, cl){
    cv <- qbinom(1-cl, sp, nv, lower.tail = FALSE) + 1
    pow <- 1 - pbinom(cv - 1, sp, av)
  }
  
  # Here is a function within the function to calculate sample size for
  # desired power.
  
  get_samp_size <- function (nv, av, sp, cl){
    n <- 10
    temp_power <- 0
    
    while (temp_power < sp){
      temp_power <- get_power(nv, av, n, cl)
      n <- n+1
    }
    
    return(n-1)
  }
  
  if (samp_size)
    result <- get_power(null_value, alt_value, samp_pow, conf_lev)
  else
    result <- get_samp_size(null_value, alt_value, samp_pow, conf_lev)
  
  return(result)
}

# Let's try calling this function with a couple sample sizes.

binom_up_pow(.5, .8, 40)
binom_up_pow(.5, .6, 40)

# Let's try calling this function for desired power.

binom_up_pow(.5, .8, .80, samp_size = FALSE)
binom_up_pow(.5, .6, .80, samp_size = FALSE)

# Now a quick check.

binom_up_pow(.5, .6, 158)

```


### Sourcing Functions

Let's clean up our code by copying the above function to a separate script
file. We'll access this function by using the *source()* function at the top
of our main script or notebook.


### Special Functions

Special functions have a different form than the usual function name and
arguments format. Base R has quite a few special functions, such as arithmetic
operators, but users can also create special functions. Types of special
functions include *infix* operators and replacement functions.

```{r}

# Most functions in R are prefix operators, meaning that the function name
# precedes the arguments. We can create infix operators by enclosing in %%.

# Let's create an infix operator that pastes two strings together.

'%+%' <- function (a, b) {
  
  new_string <- paste(a, b)
  return(new_string)
}

# Now let's try it.

"Abraham" %+% "Lincoln"

# Note that infix functions can be called just like prefix functions.

'%+%'("Abraham", "Lincoln")

# This is true even with built-in infix functions.

'*'(2, 4)

# Replacement functions are used to modify a value within an object.
# Here's a function to replace the first two values of a vector with
# some number that we hand to the function.  I don't know why we need this,
# other than to demonstrate replacement functions!

'change_first_second<-' <- function(x, value) {
  x[1:2] <- value
  return(x)
}

# Let's create a vector and call the function. Note that the call is
# a bit different.

my_vector <- 50:60
my_vector

change_first_second(my_vector) <- 20
my_vector

# Remember that everything that happens in R is a function. What we just
# learned is that when we replace values through subsetting, we are just
# using a base R replacement function!

my_vector[1] <- 10
my_vector

```


### Function Error Checking

We are well aware from previous experiences that we can make errors when we
call a function. It should be no surprise that we can also make errors when we
call our own functions. If we share our functions with others, then others can
make errors when they call our functions. Here are some examples and
considerations for anticipating that users will make errors.

```{r}

# Let's create a function with a purposeful error built in. This function
# will add five to any number less than 10. If the number is 10 or greater,
# then it will add "a", which of course is nonsense.

add_five_or_a <- function (x) {
  
  if (x < 10)
    return(x + 5)
  else
    return(x + "a")
}

# Let's call this with a number less than 10.

add_five_or_a(8)

# Now try typing this with a number greater than 10.

# add_five_or_a(12)

# Notice that we only obtained an error on the second function call. That is
# because R uses lazy evaluation.

# It is good practice to insulate users from errors. When a user provides
# a function input that is in error, we should tell them in a way that
# they will understand. That is especially true if we are going to give
# our functions to others to use in their scripts. Unfortunately, many
# R coders who share their code don't do this, so we end up with error
# messages that are difficult to interpret.

# Here's a function that will create a scatterplot and return the
# correlation. Note the error checking that can catch a couple of the
# more likely errors.

cor_plot <- function(x, y, ...){
  
  if (!(is.numeric(x) & is.numeric(y))){
    print("Use only numeric variables when calling cor_plot")
    return(NULL)
  }
  
  if (!(length(x) == length(y))) {
    print("Make sure your vectors are the same length when calling cor_plot")  
    return(NULL)
  }
  
  plot(x, y, ...)
  return(cor(x, y))
}

# Let's call it correctly first.

my_vector_1 <- c(2, 4, 6, 8, 10, 12, 14, 16)
my_vector_2 <- c(5, 1, 9, 3, 8, 7, 4, 6)

cor_plot(my_vector_1,
         my_vector_2,
         xlab = "My First Vector",
         ylab = "My Second Vector")

# Now let's create some problems.

my_vector_1 <- c("a", "b", "C")

cor_plot(my_vector_1,
         my_vector_2,
         xlab = "My First Vector",
         ylab = "My Second Vector")

my_vector_1 <- c(2, 4, 6, 8, 10, 12, 14)

cor_plot(my_vector_1,
         my_vector_2,
         xlab = "My First Vector",
         ylab = "My Second Vector")

# Let's comment out our error messages and see what would have happened.

cor_plot <- function(x, y, ...){
  
  #  if (!(is.numeric(x) & is.numeric(y))){
  #    print("Use only numeric variables when calling cor.plot")
  #    return(NULL)
  #  }
  
  #  if (!(length(x) == length(y))) {
  #    print("Make sure your vectors are the same length when calling cor.plot")  
  #    return(NULL)
  #  }
  
  plot(x, y, ...)
  return(cor(x, y))
}

my_vector_1 <- c("a", "b", "C")

# Try typing in this:

#cor_plot(my_vector_1,
#         my_vector_2,
#         xlab = "My First Vector",
#         ylab = "My Second Vector")

my_vector_1 <- c(2, 4, 6, 8, 10, 12, 14)

# Try it again:

# cor_plot(my_vector_1,
#          my_vector_2,
#          xlab = "My First Vector",
#          ylab = "My Second Vector")

# These are not as nice!

```


### Viewing Function Code

Sometimes we want to view the code for functions that we wrote to understand
why it may not be doing what we anticipated, or to consider improving or
adding to our function. We can do this without going to the source file. We
can also look at functions written by others, including Base R functions, to
see how they work and perhaps get insights for our own script writing. Another
reason to view code written by others is because we want to copy it so that we
can change or add to it.

```{r}

# We can view the code for a function by using the function name.

square_plus

# Let's look at the code for t.test

t.test

# This tells us that it is going to send all of our parameters to a particular
# t.test method. Let's see what methods there are for the t.test.

methods(t.test)

# So let's try looking at the default t.test method. Type this:

# t.test.default

# Only t.test is loaded into our environment, not t.test.default. Notice
# that it is in the stats package. Let's see if it can be exported from that
# package. Try this:

# stats::t.test.default

# Nope. So it is internal to the stats package. Let's retrieve it.

stats:::t.test.default

# We can also use the getAnywhere() function.

getAnywhere(t.test.default)

# Let's get the code for the standard deviation.

sd

# That's simple! Notice all it does is take the square root of the variance.
# Let's look at the variance function.

var

# The var function is calling a covariance function named C_cov, and it is
# using the .Call function. This tells me that it is compiled code. If I
# want to see the source, I'll have to look elsewhere.

# Let's look at an example of how we can use the source code to do what we
# want to do.

# Look at the output from the TukeyHSD function.

hsb2 <- read.csv(here("Data", "hsb2.csv"))

hsb2$ses <- factor(hsb2$ses,
                   levels = c(1:3),
                   labels = c("low", "middle", "high"))

model <- aov(hsb2$math ~ hsb2$ses)

results <- TukeyHSD(model)
plot(results)

# That's not a very good title! Let's look at the function. This is coming
# from the plot function, so I wonder if it has a special method for Tukey.

methods(plot)

# Here it is!

getAnywhere(plot.TukeyHSD)

# Let's look it over.

function (x, ...) 
{
  for (i in seq_along(x)) {
    xi <- x[[i]][, -4L, drop = FALSE]
    yvals <- nrow(xi):1L
    dev.hold()
    on.exit(dev.flush())
    plot(c(xi[, "lwr"], xi[, "upr"]), rep.int(yvals, 2L), 
         type = "n", axes = FALSE, xlab = "", ylab = "", main = NULL, 
         ...)
    axis(1, ...)
    axis(2, at = nrow(xi):1, labels = dimnames(xi)[[1L]], 
         srt = 0, ...)
    abline(h = yvals, lty = 1, lwd = 0.5, col = "lightgray")
    abline(v = 0, lty = 2, lwd = 0.5, ...)
    segments(xi[, "lwr"], yvals, xi[, "upr"], yvals, ...)
    segments(as.vector(xi), rep.int(yvals - 0.1, 3L), as.vector(xi), 
             rep.int(yvals + 0.1, 3L), ...)
    title(main = paste0(format(100 * attr(x, "conf.level"), 
                               digits = 2L), "% family-wise confidence level\n"), 
          xlab = paste("Differences in mean levels of", names(x)[i]))
    box()
    dev.flush()
    on.exit()
  }
}

# What if we edit this by putting in a title argument and them replacing
# the one that is the default.

new.plot.TukeyHSD <- function (x, new.title, ...) 
{
  for (i in seq_along(x)) {
    xi <- x[[i]][, -4L, drop = FALSE]
    yvals <- nrow(xi):1L
    dev.hold()
    on.exit(dev.flush())
    plot(c(xi[, "lwr"], xi[, "upr"]), rep.int(yvals, 2L), 
         type = "n", axes = FALSE, xlab = "", ylab = "", main = NULL, 
         ...)
    axis(1, ...)
    axis(2, at = nrow(xi):1, labels = dimnames(xi)[[1L]], 
         srt = 0, ...)
    abline(h = yvals, lty = 1, lwd = 0.5, col = "lightgray")
    abline(v = 0, lty = 2, lwd = 0.5, ...)
    segments(xi[, "lwr"], yvals, xi[, "upr"], yvals, ...)
    segments(as.vector(xi), rep.int(yvals - 0.1, 3L), as.vector(xi), 
             rep.int(yvals + 0.1, 3L), ...)
    title(main = new.title, 
          xlab = paste("Differences in mean levels of", names(x)[i]))
    box()
    dev.flush()
    on.exit()
  }
}

# Now let's try it.

new.plot.TukeyHSD(results, new.title = "Tukey Confidence Intervals")

```
