---
title: "Exercise set H"
author: "Mitchell Toomey"
date: "April 7, 2021"
output: html_notebook
---

```{r}
library(here)
library(tidyverse)
library(dplyr)
library(hrbrthemes)

```

### 1
Use the NYC flights data to construct a graph that compares carriers 
in terms of the proportion of 2013 flights that arrived on time when traveling 
from New York City to Atlanta. Your graph should include a title 
and easy-to-read labels, including the names of airlines rather than airline 
codes. You can find airline names in the Airlines Data, so you should draw 
from these data to include the names. (You completed this graph in the 
previous exercise set, but this time you should avoid 
entering airline names manually.)

```{r}

airlines <- read_csv(here("Data", "Airlines Data.csv"))

nyc_flights <- read_csv(here("Data", "NYC Flights 2013.csv"))

# Joining carriers from carrier data to flights data and renaming

flights_2 <- nyc_flights %>%
  left_join(airlines, by = "carrier")

# made a stacked proportional bar plot

flights_2 %>%
  group_by(name) %>%
  filter(dest == "ATL") %>% 
  mutate(on_off_time = (arr_delay >= 0) == TRUE | (arr_delay < 0) == FALSE) %>%
  ggplot(aes(x=name, fill=on_off_time)) + 
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Airlines", y="Proportion Arrived on Time to Atlanta", 
                fill="on time?")

# not sure how to remova NA's here
```

### 2
Construct a graph that shows how many flights each carrier had to 
Atlanta in 2013. Only include carriers that had flights to Atlanta 
but include all such carriers. For airlines with 500 or fewer flights
to Atlanta in the year, combine them into an “other” category on your graph.

```{r}

# less_than_500 <- flights_2 %>%
  #select(flight, name) %>%
  #filter(flight <= 500) %>%
  #rename(other = "flight")

# I know I need to turn all carrier's with flights less than 500 into other, 
# and then have them joined into the data and use that for the bar, but
# do not know how


#flights_2 %>%
  #group_by(name) %>%
  #filter(dest == "ATL", flight <= 500) %>%
  
  #ggplot(aes(x=name, y=flight)) + 
  #geom_bar() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  #labs(x="Airlines", y="Number of flights to Atlanta in 2013")

```

### 3
Create a tibble, UA, that only contains United Airlines flights. 
Create another tibble, AA, that only contains American Airlines flights. 
Use these tibbles and the Airports Data to determine what destinations 
United Airlines flies to from NYC that American Airlines does not, 
and vice versa.

```{r}



```

### 4
Import the Small Cities Health Data into a tibble named “health”. Construct a 
scatterplot to illustrate the relationship of the number of 
physicians (response variable) and the number of hospital beds
(explanatory variable).

```{r}

health <- read_csv(here("Data", "Small Cities Health Data.csv"))
health <- as_tibble(health)

ggplot(health, aes(x=hosp.beds, y=doctors)) + 
    geom_point() +
    labs(x="Number of Hospital Beds", y="Number of Doctors")

```

### 5
Construct a scatterplot for the same two variables, but this time only using 
cities with less than 1,000 hospital beds. Put a regression line in the plot.

```{r}
# health data set with less than 1000 hospital beds

health_2 <- filter(health, hosp.beds < 1000)

ggplot(health_2, aes(x=hosp.beds, y=doctors)) + 
    geom_point() +
    geom_smooth(method=lm , color="red", se=FALSE) + 
    labs(x="Number of Hospital Beds", y="Number of Doctors")

```

### 6
The Small Cities Densities file contains population density data for the same small cities. Add the
density variable to the health tibble

```{r}

small_city_density <- read_csv(here("Data", "Small Cities Densities.csv"))

check_density <- small_city_density %>%
  count(density)

check_density %>%
  filter(n > 1)

# I see that 7 rows have more than 1 n, so we need to figure out how to make 
# them a primary key

# see if they are accidental duplicates

small_city_density %>%
  filter((density == 57 | density == 80 | density == 86 |
            density == 95 | density == 106 | density == 108))

# no duplicate city codes, so just diff cities with same density

check_density <- small_city_density %>%
  count(city.code, density)

check_density %>%
  filter(n > 1)

small_city_density

# I didn't even realize city code was a primary key! But oh well, it got joined.

health <- health %>%
  left_join(small_city_density)

```

### 7
Add a categorical density variable to the health tibble with levels “Low” (density up to 80), “Middle”
(density from 80 to 120), and “High” (density above 120). Construct a bar graph to illustrate how
many cities there are at each density level.

```{r}
health$density_levels <- health$density

health$density_levels[health$density<=79] <- "Low"
health$density_levels[health$density>=80 & health$density<=120 ] <- "Middle"
health$density_levels[health$density>=121] <- "High"

health$density_levels <- as_factor(health$density_levels)

health %>%
  mutate(density_levels = fct_relevel(density_levels, "Low", "Middle", "High")) %>%
    ggplot(aes(x=density_levels, y=density)) + 
      geom_bar(stat = "identity") +
      scale_fill_manual(values = c("red", "green", "blue")) +
      theme(legend.position="none") +
      labs(x="Density Levels", y="Number of Cities")

```

### 8
Construct the same bar graph you constructed in the previous exercise but reverse the order of the bars in the bar graph

```{r}

nyc_flights %>%
  group_by(carrier) %>%
  filter(dest == "ATL") %>% 
  mutate(on_off_time = (arr_delay >= 0) == TRUE | (arr_delay < 0) == FALSE) %>%
  mutate(company = recode(carrier, 
       "9E" = "Endeavor Air Inc.",
       "DL" = "Delta Air Lines Inc.",
       "EV" = "ExpressJet Airlines Inc.",
       "FL" = "AirTrain Airways Corporation",
       "MQ" = "Envoy Air",
       "UA" = "United Air Lines Inc.",
       "WN" = "Southwest Airlines Co.")) %>% 
  ggplot(aes(x=fct_rev(company), fill=on_off_time)) + 
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Airlines", y="Proportion Arrived on Time to Atlanta", 
                fill="on time?")
 

```

