---
title: "Exercise D"
author: "Mitchell Toomey"
date: February 18, 2021
output: html_notebook
---

```{r}
library(here)
```

### 1

```{r}
# Loads mount juniper elementary data into a data frame

mt_juniper_data <- read.csv(here("Data", "Mount Juniper Elementary Data.csv"))

# prints either there are no missing data or there are missing data, 
# depending on which is true

mt_juniper_missing <- anyNA(mt_juniper_data$Score)

mt_juniper_missing <- ifelse(mt_juniper_missing, 
                                "There are missing data", 
                             "There are no missing data")

print(mt_juniper_missing)
```

### 2

```{r}

# Coerce grade to factor
juniper_grade <- factor(mt_juniper_data$Grade,
                        levels = c(3, 4, 5),
                        labels = c("Grade 3", "Grade 4", "Grade 5"),
                        ordered = FALSE)

anova_grade <- aov(mt_juniper_data$Score ~ juniper_grade, 
                      data = mt_juniper_data)

# Store ANOVA in a separate object with the p-value extracted

anova_results <- summary(anova_grade)[[1]][[1, 5]]

# telling R to print not statistically significant if the p-value is > .05

anova_results <- ifelse(anova_results > .05, 
                                "Not statistically significant", 
                        "statistically significant")

print(anova_results)

```

### 3

```{r}

# Adding a "results" attribute to the anova grade object

# I am using anova_grade as the object. I am not sure why exactly I use this
# and not the anova_results object-- but the latter shows attributes as NULL.
# I think because the anova_results object only has the sign. level in it.

attributes(anova_grade)$Results<- ifelse(anova_results > .05, 
                                "Not statistically significant", 
                                "statistically Significant")

```

### 4

```{r}
# Consider a population of normally distributed scores with a mean of 100 and 
# a standard deviation of15. Write a script to draw 1,000 random samples, 
# each of size 40, from this population.

norm_pop <- replicate(1000, sample(rnorm(1000, 100, 15), size = 40))

# Calculate and store each mean in a vector of means

norm_pop_means <- colMeans(norm_pop)

# Calculate the mean of the sample means

mean_of_means <- mean(norm_pop_means)

# Determine the difference between the true mean of 100 and the mean of the 
# sample means

mean_of_means - 100

# meaning there is no real difference between the true mean and the sample mean
# this is an unbiased statistic 
```

### 5 

Write a script that will count in your console pane from 0, in increments of 
10, up to 1000. Do three different versions of the script: a) with a for loop, 
b) with a while loop, and c) with a repeat and break structure

```{r}
# for loop

for(i in seq(from=0, to=1000, by=10)){
  print(i)
}

# while loop

my_number <- 0

while (my_number <= 1000) {
  print(my_number)
  my_number <- my_number + 10
}

# repeat and break loop

my_number = NULL

my_number <- 0 

repeat {
  {
  my_number <- my_number + 10  
  print(my_number)
}
  if (my_number == 1000) {
    break
  }
}
```

### 6

The secret code file contains a hidden message. Use nested for loops to reveal 
the message by pasting together the characters in every 8th position 
(i.e., positions 8, 16, 24, and 32) on all 23 rows

```{r}
secret_code <- read.csv(here("Data", "Secret Code.csv"), header = FALSE)

secret_message <- NULL

for (i in 1:23) {
  for(j in c(8, 16, 24, 32)) {
    secret_message <- paste(secret_message, secret_code[i, j], sep = "") 
  }
}

secret_message
```

### 7

Bootstrapping is a statistical method for making inferences based on repeated 
sampling from a sample of observed values. Write a script to construct a 90% 
confidence interval for the mean using bootstrapping by taking the following 
steps. (Note: there are bootstrapping functions and packages, but do not use 
these for this assignment. Instead, follow the steps below)

```{r}
# a. load the sample sata into an object. These data contain 30 scores 
# selected at random from an unknown population

sample_data <- read.csv(here("Data", "Sample Data.csv"))

sample_data <- sample_data[, 1]                        

# b. Select and store 5,000 samples of size 30 that are sampled from the 
# Sample Data with replacement. (Hint: use the sample function)

fivethous_sample <- data.frame(matrix(ncol = 500, nrow = 30))

for (i in 1:5000) {
  fivethous_sample[ , i] <- sample(sample_data, size = 30, 
                                    replace = TRUE)
}

# c. Calculate and store the mean of every sample. Collectively these 
# means will be an estimate of the sampling distribution of the mean.

sample_means <- colMeans(fivethous_sample)

# d. The 5th and 95th percentiles of the sampling distribution are the 
# boundaries of a 90% confidence interval. Calculate these and report your 
# interval. 

bounds <- quantile(sample_means, c(0.05, 0.95))

print(bounds)

# The boundaries for a 90% confidence interval are (308.43, 343.20)
```

***

**Great work!**

**80/80**
