---
title: "Exercise Set G"
author: "Mitchell Toomey"
date: "April 2, 2021"
output: html_notebook
---

```{r}
library(here)
library(tidyverse)
library(dplyr)
library(hrbrthemes)
library(MASS)
library(lattice)

college_data <- read_csv(here("Data", "College Study Data (Clean).csv"))

johnson_elem <- read_csv(here("Data", "Johnson Elementary ELA Project.csv"))

college_data
```


1. Use the College Study Data to calculate the means of the two scales by gender, by class, and by a combination of gender and class.

```{r}

# means grouped by gender

college_data %>%
  group_by(gender) %>%
  summarize(scale1_mean = mean(scale1, na.rm = TRUE),
            scale2_mean = mean(scale2, na.rm = TRUE))

# means grouped by class

college_data %>%
  group_by(class) %>%
  summarize(scale1_mean = mean(scale1, na.rm = TRUE),
            scale2_mean = mean(scale2, na.rm = TRUE))

# means groups by class x gender

college_data %>%
  group_by(class, gender) %>%
  summarize(scale1_mean = mean(scale1, na.rm = TRUE),
            scale2_mean = mean(scale2, na.rm = TRUE))

```


2. Construct side-by-side boxplots for the GPA for freshmen and seniors.

```{r}
# subset class by freshman and senior

boxplot(gpa ~ class, data = college_data, 
        subset = class %in% c("freshman", "senior"))
  
```

3.  Write a script to import and tidy the Johnson Elementary Project Data. You have done this before, but now use a single chain of pipes and make certain the data are formally tidy. (20 points)

```{r}
library(readxl)

johnson_elem <- read_excel(here("Data", "Johnson Elementary ELA Project.xlsx"), 
                           skip = 3)

johnson_elem <- rename(johnson_elem, "Room Number" = "...1")

# Fill names of room numbers, separate tardies and absences, 
# put quarters into a single quarter variable using a single chain of pipes

johnson_elem <- johnson_elem %>%
  fill("Room Number") %>%
  separate("Tardies; Absences", into = c("Tardies", "Absences")) %>% 
  gather("1st Quarter", "2nd Quarter", "3rd Quarter", "4th Quarter",
         key = "Quarter",
         value = "Score")

# check for any problems

problems(johnson_elem)

# making sure it looks okay

johnson_elem

# writing it to a new file

write_csv(johnson_elem, here("Data", "Johnson Elementary ELA Project_Cleaned2.csv"))
```


4. Write a script to provide summary statistics for the number of tardies and absences for each room.

```{r}

johnson_elem <- read_csv(here("Data", "Johnson Elementary ELA Project_Cleaned2.csv"))

johnson_elem %>%
  group_by(`Room Number`) %>%
  summarize(mean(Tardies, na.rm = TRUE),
            sd(Tardies, na.rm = TRUE),
            min(Tardies, na.rm = TRUE),
            max(Tardies, na.rm=TRUE))

johnson_elem %>%
  group_by(`Room Number`) %>%
  summarize(mean(Absences, na.rm = TRUE),
            sd(Absences, na.rm = TRUE),
            min(Absences, na.rm = TRUE),
            max(Absences, na.rm=TRUE))

```

5. Write a script to show the frequency of missing data points for each quarter by room.

```{r}

johnson_elem %>%
  group_by(`Room Number`) %>%
  summarize(count = sum(is.na(johnson_elem$Quarter)))

```

6. Construct a graph that illustrates the median rating trends across the four quarters for the fifth- grade classes. (20 points)

```{r}

johnson_elem %>%
  group_by(Quarter) %>%
  summarize(rating = median(Score, na.rm = TRUE)) %>%
  ggplot(aes(x=Quarter, y=rating)) + 
  geom_bar(stat = "identity") +
  theme_ipsum() +
  labs(x="Quarters", y="Rating")

# what Dr. Seaman did-- putting it here because I like it and want it for the 
# future

ELA_Data %>%
  group_by(Room, Quarter) %>%
  summarize(Median = median(Rating, na.rm = TRUE)) %>%
  ggplot(aes(x = Quarter, y = Median, group = Room)) +
  facet_wrap(vars(Room)) +
  geom_point() +
  geom_line() +
  labs(title = "Quarter Trends in Ratings by Room")

```

7. Use the NYC flights data to construct a graph that compares carriers in terms of the proportion of 2013 flights that arrived on time when traveling from New York City to Atlanta. Your graph should include a title and easy-to-read labels, including the names of airlines rather than airline codes. You can find airline names in the Airlines Data.
(40 points)

```{r}

library(readxl)

# read in the data 

nyc_flight <- read_excel(here("Data", "NYC Flights 2013.xlsx"))

# made a stacked proportional bar plot

nyc_flight %>%
  group_by(carrier) %>%
  filter(dest == "ATL") %>% 
  mutate(on_off_time = (arr_delay >= 0) == TRUE | (arr_delay < 0) == FALSE) %>%
  mutate(company = recode(carrier, 
       "9E" = "Endeavor Air Inc.",
       "DL" = "Delta Air Lines Inc.",
       "EV" = "ExpressJet Airlines Inc.",
       "FL" = "AirTrain Airways Corporation",
       "MQ" = "Envoy Air",
       "UA" = "United Air Lines Inc.",
       "WN" = "Southwest Airlines Co.")) %>%
  ggplot(aes(x=company, fill=on_off_time)) + 
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Airlines", y="Proportion Arrived on Time to Atlanta", 
                fill="on time?")
  
```

***

**Excellent**

**120/120**
