---
title: "Music Performance Data"
author: "Mitchell Toomey"
date: "March 15, 2021"
output: html_notebook
---

This notebook will examine the percent agreement between raters of a high school
music competition. Based on the percent agreements between raters, we will then
determine 1st place, 2nd place, and 3rd place winners for the music competition. 

```{r include=FALSE}
library(here)

library(formattable)

source(here("Functions", "Project B functions.R"))

school_rating <- read.csv(here("Data", "Music Performance Data.csv"),
                          row.names = "School")
```

#### Tables 1.1 and 1.2 examine the percent agreement and rank the winners based 
#### only on the original dataset. 

----

#### Table 1.1. Percent Agreement Among Raters with Original Data

```{r echo=FALSE}
# study the consistency of raters using
  # the original ratings

  percent_agreement <- rate_my_data(school_rating)  

  percent_agreement <- formattable(percent_agreement, 
            align = c("l","c","c","c","c", "c", "c", "c", "c"))
  
  percent_agreement
```
We see there is some consistency in mean percent agreement between all the raters,
except for rater e. The means for raters d and f are also low compared to other raters,
though not as low as rater e. 


#### Table 1.2. Ranking the Winners Based on Percent Agreement in the Original Dataset
```{r echo=FALSE}
final_decision <- final_decision(school_rating)

  final_decision <- formattable(final_decision,
                                align= c("l", "c"))

  final_decision
```
We can see here that Bedford Falls High School takes first
place, Pearl Bailey High School takes 2nd place, and Twin Peaks High School 
takes 3rd place.

-----

#### Tables 2.1 and 2.2 examine the percent agreement and rank the winners after 
#### collapsing the original rating scale into only three categories. 

----

#### Table 2.1. Percent Agreement Among Raters with Three Categories

```{r echo=FALSE}
# collapsing into three categories in groups of two

  # Collapse into 3 Categories 

  three_cats <- matrix(c(1, 2, 3, 4, 5, 6),
               nrow = 3, 
               byrow = TRUE) 

  collapse_cats_3 <- collapsed(school_rating, three_cats)

  # PA for three categories 

  percent_agreement_3 <- rate_my_data(collapse_cats_3)

  percent_agreement_3 <- formattable(percent_agreement_3, 
            align =c("l","c","c","c","c", "c", "c", "c", "c"))

  percent_agreement_3
```

We see there is more consistency in mean percent agreement between all the raters than
in the original dataset, again, except for rater e.


#### Table 2.2. Ranking the Winners Based on Percent Agreement in the Dataset with 
#### Three Collapsed Categories
```{r echo=FALSE}
# Decision for 3 collapsed categories

  three_data_decision <- final_decision(collapse_cats_3)

  three_data_decision <- formattable(three_data_decision,
                                align= c("l","c","c","c","c", "c", 
                                         "c", "c", "c"))
  three_data_decision
```

We can see from this table that the categories may be too collapsed to determine
a clear winner. 

----

#### Tables 3.1 and 3.2 examine the percent agreement and rank the winners after 
#### collapsing the original rating scale into five categories. 

----

#### Table 3.1. Percent Agreement Among Raters with Five Categories

```{r echo=FALSE}
# collapsing into five categories by collapsing somewhat weak and somewhat strong 
  # categories into a single not weak or strong category

  five_cats <- matrix(c(1, 1, 2, 2, 3, 4, 5, 5, 6, 6),
                    nrow=5,
                    byrow=TRUE)

  five_cats_collapse <- collapsed(school_rating, five_cats)

  five_cats_collapse <- formattable(five_cats_collapse, 
                             align= c("l", "c", "c", "c", "c", "c"
                                      ,"c","c", "c"))
# PA for 5 Categories 

  percent_agreement_5 <- rate_my_data(five_cats_collapse)

  percent_agreement_5 <- formattable(percent_agreement_5, 
            align =c("l","c","c","c","c", "c", "c", "c", "c"))

  percent_agreement_5
```
We see here that the percent agreement between raters looks similar to the other
two tables. Again, with rater e being significantly lower than all other raters. 

#### Table 3.2. Ranking the Winners Based on Percent Agreement in the Dataset with 
#### Five Collapsed Categories

```{r echo=FALSE}
# determine which orchestras win first, second, and third place

  five_collapse_final <- final_decision(five_cats_collapse)

  five_collapse_final <- formattable(five_collapse_final, 
                                   align= c("l", "c", "c", "c", "c", "c"
                                      ,"c","c", "c"))
  five_collapse_final

```

We can see from the five category data that Bedford Falls High School takes first place, 
Pearl Bailey High School takes 2nd place, and Twin Peaks High School takes 3rd place.
This is consistent with the original dataset. However, the interrater agreement
is much higher with five categories compared to the original dataset. 
Thus, five categories may be most appropriate for the rankings of this music competition.
