---
title: "Exercise Set F"
author: "Mitchell Toomey"
date: "March 28, 2021"
output: html_notebook
---

# 1

Prepare the College Study Data for analysis by cleaning it and importing it into R. Each variable in your data frame should be the most appropriate type of vector for the data. When you clean and import the data, you should also screen the data by checking to make sure that the data for each variable are within the specified range. If you find incorrect data and are quite certain that you know what the data should be, change the data to the correct values. If you are not sure, indicate in the cells that the data are missing.

```{r}
library(here)
library(tidyverse)

college_data <- read_csv(here("Data", "College Study Data.csv"))

college_data

college_data_probs <- problems(college_data)

college_data_probs
```

This is me comparing the original data to after the data have been read for the
rows that have problems. The read.csv function let me know what typo'd
data was actually entered so I know what to change

```{r}

college_data_original <- read.csv(here("Data", "College Study Data.csv"))
college_data_original[college_data_probs$row, ]
college_data[college_data_probs$row, ]

```

Correcting the gpa of row 11780

```{r}
college_data$gpa[11780] <- 2.80

college_data[11780, ]
```

Shifting in columns to the right for one of the rows in the previous
chunk

```{r}

college_data$id[10274] <- 10274
college_data$gender[10274] <- "male"
college_data$class[10274] <- "junior"
college_data$gpa[10274] <- 2.68
college_data$scale1[10274] <- 3
college_data$scale2[10274] <- 8

college_data[10274, ]
```

Correcting the scale two score 
```{r}
college_data$scale2[12225] <- 4

college_data[12225, ]
```

Now to manually check each variable to make sure everything is good to go

Now to check if the data are within range. First the IDs, and making sure 
each ID is unique which means they are all represented
```{r}
min(college_data$id)
max(college_data$id)

# checking that all IDs are unique

length(unique(college_data$id))
```

Now checking gender

```{r}
table(college_data$gender)
```

I will change the capital F Female to lowercase f female

```{r}
# looking for Female

college_data$id[college_data$gender == "Female"]

# checking the row

college_data[5513, ]

# changing it

college_data$gender[5513] <- "female"

# checking again

college_data[5513, ]
```

Now to change capital M Male to lowercase m male

```{r}
# looking for Male

college_data$id[college_data$gender == "Male"]

# checking the row

college_data[8963, ]

# changing it

college_data$gender[8963] <- "male"

# checking again

college_data[8963, ]

```

Now to change the senior title from the gender column

```{r}
# looking for senior

college_data$id[college_data$gender == "senior"]

# checking the row

college_data[3020, ]

# gender and class were swapped!
# changing gender to female

college_data$gender[3020] <- "female"

# changing class to senior

college_data$class[3020] <- "senior"


# checking again

college_data[3020, ]
```

Checking gender again

```{r}
table(college_data$gender)
```

Now checking the class variable

```{r}
table(college_data$class)
```

Finding and correcting junirr to junior

```{r}
# looking for junirr

college_data$id[college_data$class == "junirr"]

# checking the row

college_data[11928, ]

# changing class to senior

college_data$class[11928] <- "junior"


# checking again

college_data[11928, ]
```

Now to find SENIOR and change it to senior

```{r}
# looking for SENIOR

college_data$id[college_data$class == "SENIOR"]

# checking the row

college_data[6472, ]

# changing class to senior

college_data$class[6472] <- "senior"


# checking again

college_data[6472, ]
```

Rechecking class to make sure it is okay

```{r}
table(college_data$class)
```

Now checking gpa

```{r}
min(college_data$gpa, na.rm = TRUE)
max(college_data$gpa, na.rm = TRUE)
```

Data should only max our at 4.00. I need to find GPAs above 4.00

```{r}
college_data$id[college_data$gpa > 4.00 & !is.na(college_data$gpa)]

# looking at the row

college_data[10918, ]
```

Not sure what that 12.96 is supposed to be so just gonna make it NA

```{r}
college_data$gpa[10918] <- NA

college_data[10918, ]
```

Checking to make sure gpa all checks out

```{r}
min(college_data$gpa, na.rm = TRUE)
max(college_data$gpa, na.rm = TRUE)
```

Now to check scale1

```{r}
table(college_data$scale1)
```

Someone put -99 as missing values so I need to make them NA

```{r}
college_data$scale1[college_data$scale1 == -99] <- NA
```

Now to check scale1 again to see if -99's have been changed

```{r}
table(college_data$scale1)
```

Now to check scale2

```{r}
table(college_data$scale2)
```

Okay, gotta change the -999, -99, -9, and 12 to NA

```{r}
college_data$scale2[college_data$scale2 == -999] <- NA

college_data$scale2[college_data$scale2 == -99] <- NA

college_data$scale2[college_data$scale2 == -9] <- NA

college_data$scale2[college_data$scale2 == 12] <- NA

```

Now to check scale2 again

```{r}
table(college_data$scale2)
```

Read_csv may have changed the classes depending on the errors, so I want
to check the classes

```{r}
college_data
```

id, gender should be categorical cus their values do not have numerical meaning
make scale1 and scale2 to integers because it will be prettier since they are 
not doubles

```{r}
college_data$id <- factor(college_data$id)
college_data$gender <- factor(college_data$gender)
college_data$class <- factor(college_data$class)
college_data$scale1 <- as.integer(college_data$scale1)
college_data$scale2 <- as.integer(college_data$scale2)
```

checking class again now to make sure

```{r}
college_data
```

I am finally ready to write the cleaned data

```{r}
# now writing the cleaned data

write.csv(college_data, here("Data", "College Study Data (Clean).csv"))
```


# 1 -- at end of problem one doing this per Dr. Seaman's request in tutorial sesh

```{r}
rm(list = ls())

source(here("Functions", "College Data Check.R"))

clean_collegedata <- read.csv(here("Data", "College Study Data (Clean).csv"))

College_Data_Check(clean_collegedata)
```

# 2 
Import and Tidy the Johnson Elementary Data

```{r}
library(readxl)

johnson_elem <- read_excel(here("Data", "Johnson Elementary ELA Project.xlsx"), 
                           skip = 3)

johnson_elem <- rename(johnson_elem, "Room Number" = "...1")

# Fill names of room numbers

johnson_elem <- johnson_elem %>%
  fill("Room Number")

# Separate tardies and absences

johnson_elem <- johnson_elem %>%
  separate("Tardies; Absences", into = c("Tardies", "Absences"))

# Group quarter scores into one quarter variable

johnson_elem <- johnson_elem %>%
  gather("1st Quarter", "2nd Quarter", "3rd Quarter", "4th Quarter",
         key = "Quarter",
         value = "Score")

johnson_elem

# writing it to a new file

write_csv(johnson_elem, here("Data", "Johnson Elementary ELA Project_Cleaned.csv"))

```

